TRIAD: Towards Real-time, Interactive Analysis of Data

### 2015-05-01 paper.tex
final iteration...

%% indicate IEEE Member or Student Member in form indicated below
\author{Remco Chang \and Danyel Fisher \and Jeffrey Heer \and Carlos Scheidegger}
\authorfooter{
%% insert punctuation at end of each item
\item
 Remco Chang is with Tufts University. email: remco@cs.tufts.edu
\item
 Danyel Fisher is with Microsoft Research. email: danyelf@microsoft.com
\item
 Jeffrey Heer is with the University of Washington. email: jheer@uw.edu
\item
 Carlos Scheidegger is with the University of Arizona. email: cscheid@cs.arizona.edu
}

@@

\begin{document}
\maketitle
%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.
\section{Goals}
%% the only exception to this rule is the \firstsection command
The goal of this workshop is to foster innovative work in the backends
that will support the next generation of interactive data analysis
tools. We envision such backends to require novel insights in database
technology, algorithm design, and information visualization techniques.

In the current design of database and analytics systems, interactive
data analysis concerns are usually an afterthought. Although recent
architectural changes like columnar databases have enabled faster
visualization systems, these databases are often not specified optimized for data visualization purposes but for general acceleration of queries. In
this workshop, we will explore the consequences of pushing the
concerns of interactive data analysis deep into the ``computational
fabric'' of our systems: the back-end infrastructure of data storage,
organization, and retrieval. If we design with the knowledge that
every analysis is eventually read by a human, bound by perceptual
limitations and latency constraints, what are the consequences for
database design, or data analysis and visualization algorithms? Are
there fundamental tradeoffs? What can we give up?

There have been previous workshops in VIS that have investigated
issues of large data visualization, such as LDAV (and LDAV will be a
co-located symposium this year at VIS). Our focus is markedly
different from that of LDAV. LDAV focuses on simulation and imaging
data, while we believe current techniques in information visualization
and visual analytics are close to hitting a ceiling in terms of
performance. We propose this workshop as a venue to discuss these
limits, and fundamental new ways to attack them.

\section{Technical Scope}
The scope of this workshop will span, among possibly other fields,
information visualization, visual analytics, algorithm design,
database architecture, and statistics. Although this is clearly a
broad scope, the workshop will focus on the backends for interactive
tools. We have made significant progress in our understanding of
perceptual and human-interaction issues behind interactive
visualization design. However, much of this understanding has not, in
general, been translated into new designs for our backend
infrastructure. We think now is the right time to have a discussion of
backend design for interactive data analysis. Following the principle
of ``No One Size Fits All'' in designing databases~\citeref{stonebraker2005one}, recent
publications by the organizers have shown that by taking advantage of
needs and limitations of interactive visualization systems, databases
can be optimized by making tradeoffs in novel and interesting
ways. For example, imMens~\citeorgref{Liu:2013:IRT}, profiler~\citeorgref{Kandel:2012:PIS}, and nanocubes~\citeorgref{Lins:2013:NFR} have
shown that careful algorithm design can enable extremely fast
interactive visualizations of massive data. Scalr~\citeorgref{Battle:2013:DRO} has shown that
traditional query planners can fruitfully take advantage of perceptual
and resolution limits. ForeCache~\citeorgref{Battle:2015:VLDB} utilizes a predictive prefetching
strategy by learning a user’s interaction patterns and prefetching
data from server to support real-time visual exploration. Lastly,
SampleAction~\citeorgref{Fisher:2012:TMI} takes advantage of streaming databases and
incorporates HCI principles that users need immediate feedback, but
are willing to wait for high-fidelity results.

In addition to the work by the organizers, researchers in the database
community have also made significant stride in developing databases
specifically for interactive visualization, analysis, and exploration
of data. For example, systems like Tableau utilizes columnar databases
which provide speedups for typical visual analytics tasks that require
operations across columns (instead of the traditional rows). BlinkDB~\citeref{Agarwal:2013:BQW}
uses a sampling strategy to build a smaller, but representative
database from a larger one. Map-D~\citeref{mapd} takes advantage of modern
graphics card capabilities and resulted in a specialized database that
runs entirely within (an array of) graphics cards. Finally, SciDB~\citeref{cudre2009demonstration}
breaks away from the traditional relational data structure and is
designed to store and query multidimensional scientific data that are
stored in an array-based structure.

Given the increasing number of publications and backend systems that
can be tailored or applied to interactive visual exploration of large
data, we believe that it is very important for the InfoVis and the
VAST communities to begin having discussions and to develop a research
agenda that focuses on the backend technologies. In this proposed
workshop, we seek to invite members from the database community
(including authors of the aforementioned papers) to meet with the VIS
community and begin such dialogs. As many of these database
researchers are current collaborators of the organizers (see Section~\ref{sec:invitees} below for a list of intended invitees), we believe that the
organizing team is well suited to influence research not only in
information visualization, but in these other related fields as well.

\section{Format, Length, and Planned Activities}
We propose a half-day workshop that focuses on discussion between the
VIS and database researchers. The workshop will solicit submissions of
position papers as well as short research papers from both the VIS and
database communities. These papers will be reviewed by the organizing
team. The authors of the selected top papers will be asked to each
give a short presentation of their work (or position).

The goal of the presentation is to allow the attendees to get an
overview of the different perspectives and approaches of data storage,
organization, and retrieval techniques that can be tailored
specifically to support interactive data visualization and
exploration. We anticipate that these presentations will take up
approximately the first half of the workshop.

@@

It is relevant to note that the organizing team is also currently
seeking a followup ``closed-door'' (invitation only) meeting that will
take place in the afternoon outside of the main VIS venue. The purpose
of this closed-door meeting is to allow for a more intimate discussion
between enthusiastic researchers who are actively seeking
collaborations across the aisle. Although it is not the goal of the
organizing team to be exclusive, it is our belief that a more focused
discussion will have a higher likelihood towards developing a cohesive
research agenda. Note that this is not an extraordinary proposal: such
a format has been successfully employed by other workshops in previous
years at VIS, most notably by the BELIV workshop~\citeref{beliv}.

@@

\subsection*{Early or mid-career researchers}
\begin{itemize}[topsep=0pt, partopsep=0pt, itemsep=-3pt]
\item Eugene Wu (Columbia)
\item Aaron Elmore (U. Chicago)
\item Joey Gonzalez (Berkeley AMPLab, CMU, GraphLab/Dato)
\item Bill Howe (UW)
\item Arnab Nandi (OSU)
\item Cecilia Aragon (UW; astronomy/scivis background)
\item Jenny Duggan (Northwestern)
\item Mike Cafarella (Michigan)
\end{itemize}

\subsection*{Junior Researchers (Students, etc.)}
\begin{itemize}[topsep=0pt, partopsep=0pt, itemsep=-3pt]
\item Manasi Vartak (MIT, SeeDB)
\item Todd Mostak (MapD)
\item Leilani Battle (MIT, Tile Prefetching)
\item Dominik Moritz (UW)
\item Holger Pirk (MIT postdoc, databases)
\end{itemize}

@@

We should ask for a setup split halfway with seats for lectures and space for posters in the back. TBD.

@@

TBD.

>>>>

%% indicate IEEE Member or Student Member in form indicated below
\author{Remco Chang\\
	\scriptsize Tufts University 
	\and Danyel Fisher\\
	\scriptsize Microsoft Research 
	\and Jeffrey Heer\\
	\scriptsize University of Washington
 	\and Carlos Scheidegger\\
 	\scriptsize University of Arizona}
\authorfooter{
%% insert punctuation at end of each item
\item
 Remco Chang is with Tufts University. email: remco@cs.tufts.edu
\item
 Danyel Fisher is with Microsoft Research. email: danyelf@microsoft.com
\item
 Jeffrey Heer is with the University of Washington. email: jheer@uw.edu
\item
 Carlos Scheidegger is with the University of Arizona. email: cscheid@cs.arizona.edu
}

@@

\begin{document}
\maketitle
%% The ``\maketitle'' command must be the first command after the
%% ``\begin{document}'' command. It prepares and prints the title block.
\section{Goals}
%% the only exception to this rule is the \firstsection command
The goal of this workshop is to foster innovative work in the backend that will support the next generation of interactive data analysis tools. 
We envision such backends to require novel insights in database technology, algorithm design, and information visualization techniques.

In the current design of database and analytics systems, interactive data visualization and analysis concerns are usually an afterthought. 
Although recent architectural changes like columnar databases have enabled faster visualization systems, these databases are often not specifically optimized for data visualization purposes but for general acceleration of queries. 
In this workshop, we will explore the consequences of pushing the concerns of interactive data visualization and analysis deep into the ``computational fabric'' of our systems: the back-end infrastructure of data storage, organization, and retrieval. 
If we design with the knowledge that every analysis is eventually read by a human, bound by perceptual limitations and latency constraints, what are the consequences for database design, or data analysis and visualization algorithms? 
Are there fundamental tradeoffs? 
What can we give up?

There have been previous workshops in VIS that have investigated issues of large data visualization, such as LDAV (and LDAV will be a co-located symposium this year at VIS). 
Our focus is markedly different from that of LDAV. 
While LDAV focuses on simulation and imaging data, we believe current backend techniques in information visualization and visual analytics are close to hitting a ceiling in terms of performance. 
We propose this workshop as a venue to discuss these limits, and fundamental new ways to attack them.

\section{Technical Scope}
The scope of this workshop will span, among possibly other fields, information visualization, visual analytics, algorithm design, database architecture, and statistics. 
Although this is clearly a broad scope, the workshop will focus on backends for interactive tools. 
As a community, visualization researchers have made significant progress in our understanding of perceptual and human-interaction issues behind interactive visualization design. 
However, much of this understanding has not, in general, been translated into new designs for our backend infrastructure. 
We think now is the right time to have a discussion of backend design for interactive data analysis. 
Following the principle of ``No One Size Fits All'' in designing databases~\citeref{stonebraker2005one}, recent publications by the organizers have shown that by taking advantage of the needs and limitations of interactive visualization systems, databases can be optimized by making tradeoffs in novel and interesting
ways. 
For example, imMens~\citeorgref{Liu:2013:IRT}, profiler~\citeorgref{Kandel:2012:PIS}, and nanocubes~\citeorgref{Lins:2013:NFR} have shown that careful algorithm design can enable extremely fast interactive visualizations of massive data. 
Scalr~\citeorgref{Battle:2013:DRO} has shown that traditional query planners can fruitfully take advantage of perceptual and resolution limits. 
ForeCache~\citeorgref{Battle:2015:VLDB} utilizes a predictive prefetching strategy by learning a user's interaction patterns and prefetching data from the server to support real-time visual data exploration. 
Lastly, SampleAction~\citeorgref{Fisher:2012:TMI} takes advantage of streaming databases and
incorporates the HCI principle that users need immediate feedback, but are willing to wait for high-fidelity results.

In addition to the work by the organizers, researchers in the database community have also made significant strides in developing databases specifically for interactive visualization, analysis, and exploration
of data. 
For example, systems like Tableau~\citeref{stolte2002polaris} utilize columnar databases which provide speedups for typical visual analytics tasks that require operations across columns (instead of the traditional rows). 
BlinkDB~\citeref{Agarwal:2013:BQW} uses a sampling strategy to build a smaller, but representative database from a larger one. 
MapD~\citeref{mapd} takes advantage of modern graphics card capabilities and resulted in a specialized database that runs entirely within (an array of) graphics cards. 
Finally, SciDB~\citeref{cudre2009demonstration} breaks away from the traditional relational data structure and is designed to efficiently store and query multidimensional scientific data that are stored in an array-based structure.

Given the increasing number of publications and backend systems that can be tailored or applied to interactive visual exploration of large data, we believe that it is very important for the InfoVis and the
VAST communities to begin having discussions and to develop a research agenda that focuses on backend technologies. 
In this proposed workshop, we seek to invite members from the database community (including authors of the aforementioned papers) to meet with the VIS community and begin such dialogs. 
As many of these database researchers are current collaborators of the organizers (see Section~\ref{sec:invitees} below for a list of intended invitees), we believe that the organizing team is well suited to influence research not only in information visualization, but in these other related fields as well.

\section{Format, Length, and Planned Activities}
We propose a half-day workshop that focuses on generating discussions between VIS and database researchers. 
The workshop will solicit submissions of position papers as well as short research papers from both the VIS and database communities. 
These papers will be reviewed by the organizing team. 
The authors of the selected top papers will be invited to each give a short presentation of their work (or position).

The goal of the presentation is to allow the attendees to gain an overview of the different perspectives and approaches of data storage, organization, and retrieval techniques that can be tailored specifically to support interactive data visualization and exploration. 
We anticipate that these presentations will take up approximately the first half of the workshop.

@@

It is relevant to note that the organizing team is also currently seeking a follow-up ``closed-door'' (invitation only) meeting that will take place in the afternoon outside of the main VIS venue. 
The purpose of this closed-door meeting is to allow for a more intimate discussion between enthusiastic researchers who are actively seeking collaborations across the aisle. 
Although it is not the goal of the organizing team to be exclusive, it is our belief that a more focused
discussion will have a higher likelihood towards developing a cohesive research agenda. 
Note that this is not an extraordinary proposal: such an off-site format has been successfully employed by other workshops in previous years at VIS, most notably by the BELIV workshop~\citeref{beliv}.

@@

\subsection*{Early or mid-career researchers}
\begin{itemize}[topsep=0pt, partopsep=0pt, itemsep=-3pt]
\item Eugene Wu (Columbia)
\item Aaron Elmore (U. Chicago)
\item Joey Gonzalez (Berkeley AMPLab, CMU, GraphLab/Dato)
\item Bill Howe (UW)
\item Arnab Nandi (OSU)
\item Cecilia Aragon (UW)
\item Jenny Duggan (Northwestern)
\item Mike Cafarella (Michigan)
\end{itemize}

\subsection*{Junior researchers (students, etc.)}
\begin{itemize}[topsep=0pt, partopsep=0pt, itemsep=-3pt]
\item Manasi Vartak (MIT)
\item Todd Mostak (MapD)
\item Leilani Battle (MIT)
\item Dominik Moritz (UW)
\item Holger Pirk (MIT postdoc)
\end{itemize}

@@

In order to facilitate both paper presentations and open discussions, we seek to have a room that has space for both seats and for posters.
A podium with audio/video support will be necessary for the paper presentations.
In addition, some supplies for sketching and drawing would be beneficial in facilitating group discussions (e.g. big Easel Pads with markers).

@@

There are three goals for this workshop. 
First, we hope to engage both researchers in the VIS and the database communities.
Should we be successful in attracting and inviting database researchers, the workshop can serve as a forum for true interdisciplinary discussion and collaboration.

Second, as we will accept both position papers and (short) research papers, we aim to compile the accepted papers into a short proceeding that we will disseminate to the workshop attendees. 
While we do not anticipate this proceeding to contain cutting-edge research contributions, we do envision the content to serve as a starting point for an open dialog between the communities.

Lastly, depending on the outcomes of these discussions, we aim to sketch out an initial outline for a research agenda for this burgeoning field.
This research agenda will summarize the current state-of-the-art in both the VIS and database communities concerning backend supports for interactive data visualization and analysis.
In addition, it will highlight the top challenges that need to be addressed that will hopefully spur future research in this field.
The four organizers will enlist attendees of the workshop to help author this research agenda with the hope that the resulting paper will be published in a venue that is visible to both the VIS and database communities.
